```{r global_options, include=FALSE}
library(knitr)
opts_chunk$set(fig.align="center", fig.height=3, fig.width=4)
```
## In-class worksheet 5
**Jan 30, 2018**

## 1. Tidy data

Is the `iris` dataset tidy? Explain why or why not.
```{r}
head(iris)
```
*Yes, it is tidy. It meets the three rules*

Is the `HairEyeColor` dataset tidy? Explain why or why not.
```{r}
HairEyeColor
```
*No, it is not. The variablea are Eye color, Hair color, Sex, and person.*

## 2. Selecting rows and columns
All subsequent code will be based on the dplyr library, which is part of the tidyverse. So we first have to load this library:
```{r message=FALSE}
library(tidyverse)
```

Now, using the dplyr function `filter()`, pick all the rows in the `iris` dataset that pertain to species setosa, and store them in a new table called `iris.setosa`.
```{r }
iris.setosa <- filter(iris, Species=='setosa')
head(iris.setosa)
```

Pick all the rows in the `iris` dataset where species virginica has a sepal length > 7.
```{r }
filter(iris, Sepal.Length>7 & Species=='virginica')
```

Are there any cases in the `iris` dataset for which the ratio of sepal length to sepal width exceeds the ratio of petal length to petal width? Use `filter()` to find out.
```{r }
filter(iris, Sepal.Length/Sepal.Width > Petal.Length/Petal.Width)
```


Create a pared-down table which contains only data for species setosa and which only has the columns `Sepal.Length` and `Sepal.Width`. Store the result in a table called `iris.pared`.
```{r }
iris.pared0 <- select(iris.setosa,Sepal.Length,Sepal.Width)
head(iris.pared0)

# first extract all data for species setosa
iris.setosa <- filter(iris, Species=='setosa')
# now select sepal length and width
iris.pared <- select(iris.setosa, Sepal.Length, Sepal.Width)
head(iris.pared)
```

## 3. Creating new data, arranging

Using the function `mutate()`, create a new data column that holds the ratio of sepal length to sepal width. Store the resulting table in a variable called `iris.ratio`.
```{r }
iris.ratio <- mutate(iris, sepal.length.to.width = Sepal.Length/Sepal.Width)
head(iris.ratio)
```

Order the `iris.ratio` table by species name and by increasing values of sepal length-to-width ratio.
```{r}
iris.ratio.ordered <- arrange(iris.ratio, Species, sepal.length.to.width)
head(iris.ratio.ordered)
```

## 4. Grouping and summarizing

Calculate the mean and standard deviation of the sepal lengths for each species. Do this by first creating a table grouped by species, which you call `iris.grouped`. Then run `summarize()` on that table.
```{r }
iris.grouped <- group_by(iris, Species)
head(iris.grouped)

summarize(iris.grouped,
            mean.sepal.length = mean(Sepal.Length),
            sd.sepal.length = sd(Sepal.Length))
```

Use the function `n()` to count the number of observations for each species.
```{r }
summarize(iris.grouped, count = n())
```

For each species, calculate the percentage of cases with sepal length > 5.5.
```{r }
summarize(iris.grouped, percent = sum(Sepal.Length>5.5)/n())
```


## 5. If this was easy

Take the `iris.ratio` data set you have created and plot the distribution of sepal length-to-width ratios for the three species.
```{r }
# ggplot2 is part of tidyverse, so we don't need to load it separately
theme_set(theme_bw(base_size=12)) # change the ggplot2 theme
ggplot(iris.ratio, aes(x=sepal.length.to.width, fill=Species)) + geom_density(alpha=.3)
```

Now plot sepal length-to-width ratios vs. sepal lengths. Does it look like there is a relationship between the length-to-width ratios and the lengths? Does it matter whether you consider each species individually or all together? How could you find out?
```{r }
ggplot(iris.ratio, aes(y=sepal.length.to.width, x=Sepal.Length, color=Species)) + geom_point()
```

```{r }
# Setosa
setosa.ratio <- filter(iris.ratio, Species=='setosa')
cor.test(setosa.ratio$Sepal.Length, setosa.ratio$sepal.length.to.width)
```

```{r }
# Versicolor
versicolor.ratio <- filter(iris.ratio, Species=='versicolor')
cor.test(versicolor.ratio$Sepal.Length, versicolor.ratio$sepal.length.to.width)
```

```{r }
# Virginica
virginica.ratio <- filter(iris.ratio, Species=='virginica') 
cor.test(virginica.ratio$Sepal.Length, virginica.ratio$sepal.length.to.width)
```
```
