```{r global_options, include=FALSE}
library(knitr)
opts_chunk$set(fig.align="center", fig.height=3, fig.width=4)
```
## In-class worksheet 7
**Feb 6, 2018**
*Yao in the class*

In this worksheet, we will use the libraries tidyverse and nycflights13:
```{r message=FALSE}
library(tidyverse)
theme_set(theme_bw(base_size=12)) # set default ggplot2 theme
library(nycflights13)
```
The nycflights13 package contains information about all planes departing fron New York City in 2013. 

## 1. Joining tables

The following two tables list the population size and area (in sq miles) of three major Texas cities each:
```{r}
population <- read.csv(text=
"city,year,population
Houston,2014,2239558
San Antonio,2014,1436697
Austin,2014,912791
Austin,2010,790390", stringsAsFactors = FALSE)
population
area <- read.csv(text=
"city,area
Houston,607.5
Dallas,385.6
Austin,307.2", stringsAsFactors = FALSE)
area
```
*Note: We can read data from .txt/.csv with "stringsAsFactors = FALSE".*

Combine these two tables using the functions `left_join()`, `right_join()`, and `inner_join()`. How do these join functions differ in their results?

```{r}
left_join(population,area) # keep the left data, add variables from the right data
right_join(population,area) # keep the right data, add variables from the right data
inner_join(population,area) # Intersection observations for column 1, all variables
semi_join(population,area) # Intersection observations for column 1, left variables
full_join(population,area) # All observations, all variables, with NA filled
anti_join(population,area) # observations for column 1 in left but not in right, left variables
```

## 2. Relationship between arrival delay and age of plane

The table `flights` from nycflights13 contains information about individual departures:
```{r}
flights
```
Individual planes are indicated by their tail number (`tailnum` in the table). Calculate the mean arrival delay (`arr_delay`) for each tail number. Do you notice anything unusual in the result? Try to calculate the mean with and without adding the option `na.rm=TRUE`.
```{r}
flights %>% group_by(tailnum) %>% summarize(mean_arrival_delay = mean(arr_delay))
flights %>% group_by(tailnum) %>% summarize(mean_arrival_delay = mean(arr_delay,na.rm=TRUE)) -> MeanArrivalDelay# a logical value indicating whether NA values should be stripped before the computation proceeds.
```

Information about individual planes is availabe in the table `planes`:
```{r}
planes
```

In particular, this table lists the year each individual plane was manufactured. Make a combined table that holds tail number, mean arrival delay, and year of manufacture for each plane. Then plot mean arrival delay vs. year of manufacture.
```{r}
full_join(MeanArrivalDelay,planes) %>% select(tailnum,mean_arrival_delay,year) -> temp
# left_join(MeanArrivalDelay,planes) %>% select(tailnum,mean_arrival_delay,year) -> temp
temp
ggplot(temp,aes(y = mean_arrival_delay,x = year)) + geom_point()
```

## 3. Relationship between arrival delay and temperature

Now calculate the mean arrival delay for each day of the year, and store in a variable called `daily_delays`.
```{r}
flights %>% group_by(year,month,day) %>% summarize(mean_arrival_delay = mean(arr_delay,na.rm=TRUE)) -> daily_delays
daily_delays
```

We want to correlate these delay values with the temperature of each day. The data frame `weather` holds temperature measurements for each hour of each day:
```{r}
weather
```

First, calculate the mean temperature for each day, and store in a variable called `mean_temp`:
```{r}
weather %>% group_by(year,month,day) %>% summarize(mean_temp = mean(temp)) -> mean_temp
mean_temp
```

Now combine the mean delay and the mean temperature into one table, and then plot mean delay vs. mean temperature.
```{r}
full_join(daily_delays,mean_temp) %>% ggplot(aes(y = mean_arrival_delay,x = mean_temp)) + geom_point()

```

## 4. If this was easy

Find out for how many tail numbers in the `flights` data set we have no information in the `planes` data set. What do we have to pay attention to when joining the `flights` and `planes` tables?
```{r}
left_join(flights,planes) -> joined_data
flights %>% left_join(planes, by='tailnum') %>% filter(is.na(type)) %>% tally()
```
*There are 52606 such flights. It is important here to tell the left_join() function to join by tailnum, otherwise it tries to join by tailnum and year, but year has different meanings in the two tables.*

Calculate the mean arrival delay by plane model and by plane engine. Sort in order of descending mean delay. Remove all tailnumbers for which no plane information is available.
```{r}
# 1. plane model
# we first join the fligths table and the planes table to make a new table holding plane model, engine, and arrival delay
flights %>% left_join(planes, by='tailnum') %>% 
  filter(!is.na(type)) %>% 
  select(model, engine, arr_delay) -> delay_table
delay_table
# we next calculate the mean delay per model
delay_table %>% group_by(model) %>% 
  summarize(mean_delay=mean(arr_delay, na.rm=T)) %>%
  arrange(desc(mean_delay)) -> model_delay
model_delay
# 2. plane engine
# we go back to the delay_table we created above, and now calculate the mean per engine
delay_table %>% group_by(engine) %>% 
  summarize(mean_delay=mean(arr_delay, na.rm=T)) %>%
  arrange(desc(mean_delay)) -> engine_delay
engine_delay
```

