```{r global_options, include=FALSE}
library(knitr)
opts_chunk$set(fig.align="center", fig.height=3, fig.width=4)
library(ggplot2)
theme_set(theme_bw(base_size=12))
library(dplyr)
```
## In-class worksheet 11
**Feb 20, 2018**

In this worksheet, we will use the library tidyverse and ggthemes (for colorblind-friendly color scale `scale_color_colorblind()`):
```{r message=FALSE}
library(tidyverse)
theme_set(theme_bw(base_size=12)) # set default ggplot2 theme
# install.packages("ggthemes")
library(ggthemes)
```


# 1. Fitting a logistic regression model to the iris data set

We will work with the `iris` data set. Specifically, with a subset of the data that consists only of the species virginica and versicolor:
```{r}
# make a reduced iris data set that only contains virginica and versicolor species
iris.small <- filter(iris, Species %in% c("virginica", "versicolor"))
head(iris.small)
```

Fit a logistic regression model to the `iris.small` data set. Then successively remove predictors until only predictors with a p value less than 0.1 remain.
```{r}
glm.out <- glm(Species ~ Sepal.Length + Sepal.Width + Petal.Length + Petal.Width,
               data = iris.small,
               family = binomial)
summary(glm.out)
# Remove Sepal.Length because its p-value is 0.30
glm.out <- glm(Species ~ Sepal.Width + Petal.Length + Petal.Width,
               data = iris.small,
               family = binomial)
summary(glm.out)
```

Make a plot of the fitted probability as a function of the linear predictor, colored by species identity. Hint: you will have to make a new data frame combining data from the fitted model with data from the `iris.small` data frame.

```{r}
iris.predicted <- data.frame(iris.small,prob = glm.out$fitted.values,predictor = glm.out$linear.predictors)
head(iris.predicted)
iris.predicted %>% ggplot(aes(x = predictor, y = prob, color=Species)) + 
  geom_point() +
  scale_color_colorblind()
# plot the line
virg_data <- filter(iris.predicted, Species=='virginica')
vers_data <- filter(iris.predicted, Species=='versicolor')
```

Make a density plot that shows how the two species are separated by the linear predictor.
```{r}
iris.predicted %>% ggplot(aes(x = predictor, fill=Species)) +
  geom_density(alpha=.5) +
  scale_color_colorblind() #? color different?
```

# 2. Predicting the species

Assume you have obtained samples from three plants, with measurements as listed below. Predict the likelihood that each of these plants belongs to the species virginica.
```{r}
plant1 <- data.frame(Sepal.Length=6.4, Sepal.Width=2.8, Petal.Length=4.6, Petal.Width=1.8)
plant2 <- data.frame(Sepal.Length=6.3, Sepal.Width=2.5, Petal.Length=4.1, Petal.Width=1.7)
plant3 <- data.frame(Sepal.Length=6.7, Sepal.Width=3.3, Petal.Length=5.2, Petal.Width=2.3)
```

```{r}
predict(glm.out,plant1, type = 'response')
predict(glm.out,plant2, type = 'response')
predict(glm.out,plant3, type = 'response')
```


# 3. If this was easy

Pick a cutoff predictor value at which you would decide that a specimen belongs to virginica rather than versicolor. Calculate how many virginicas you call correctly and how many incorrectly given that choice.

```{r}
# how to pick the best cutoff predictor value? 0 seems good!
iris.predicted %>% filter(Species=='virginica' & predictor > 0) %>%
  select(prob) %>%
  summarize(correct_virginica=n())
iris.predicted %>% filter(Species=='virginica') %>%
  summarize(total_virginica=n())
```

Now do the same calculation for versicolor.
```{r}
iris.predicted %>% filter(Species=='versicolor' & predictor < 0) %>%
  select(prob) %>%
  summarize(correct_versicolor=n())
iris.predicted %>% filter(Species=='versicolor') %>%
  summarize(total_versicolor=n())
```

If we define a call of virginica as a positive and a call of versicolor as a negative, what are the true positive rate (sensitivity, true positives divided by all possible positives) and the true negative rate (specificity, true negatives divided by all possible negatives) in your analysis?
```{r}
# true positive rate
(49)/50*100
# true negative rate
(48)/50*100
```
