```{r global_options, include=FALSE}
library(knitr)
opts_chunk$set(fig.align="center", fig.height=3, fig.width=4)
```
## PCA: test on repeated measurements
**Mar 9, 2018**

In this worksheet, we will use the libraries ggplot2, cowplot, dplyr, and grid:
```{r message=FALSE}
library(tidyverse)
theme_set(theme_bw(base_size=12)) # set default ggplot2 theme
library(egg) # required to arrange plots side-by-side
library(grid) # required to draw arrows
library(plyr); library(dplyr)
```


# 1. PCA of the iris data set

The `iris` dataset has four measurements per observational unit (iris plant):
```{r}
head(iris)
ddply(iris,.(Species),summarize,tol_in_group=length(Species))

# repeated2: make each observation repeated twice
iris.selected <- iris[1:150,]
iris.repeated2 <- iris.selected[rep(seq_len(nrow(iris.selected)), each=2),] # each is repeated twice
head(iris.repeated2)

# repeated3: make versicolor observations repeated for 6 times
iris.selected <- iris[50:100,]
temp <- iris.selected[rep(seq_len(nrow(iris.selected)), each=5),] # each is repeated twice
iris.repeated3 <- rbind(temp,iris) # each is repeated for three times
head(iris.repeated3)
```


The basic steps in PCA are to (i) prepare a data frame that holds only the numerical columns of interest, (ii) scale the data to 0 mean and unit variance, and (iii) do the PCA with the function `prcomp()`:
```{r}
iris %>% select(-Species) %>% # remove Species column
  scale() %>%                 # scale to 0 mean and unit variance
  prcomp() ->                 # do PCA
  pca                         # store result as `pca`

# iris.repeated2
iris.repeated2 %>% select(-Species) %>% # remove Species column
  scale() %>%                 # scale to 0 mean and unit variance
  prcomp() ->                 # do PCA
  pca1                         # store result as `pca`

# iris.repeated3
iris.repeated3 %>% select(-Species) %>% # remove Species column
  scale() %>%                 # scale to 0 mean and unit variance
  prcomp() ->                 # do PCA
  pca2                        # store result as `pca`

# now display the results from the PCA analysis
pca
pca1 # iris.repeated2
pca2 # iris.repeated3
```
The main results from PCA are the standard deviations and the rotation matrix. We will talk about them below. First, however, let's plot the data in the principal components. Specifically, we will plot PC2 vs. PC1. The rotated data are available as `pca$x`:
```{r}
head(pca$x)
head(pca1$x)# iris.repeated2
head(pca2$x)# iris.repeated3
```
As we can see, these data don't tell us to which species which observation belongs. We have to add the species information back in:
```{r}
# add species information back into PCA data
pca_data <- data.frame(pca$x, Species=iris$Species)
head(pca_data)

pca_data1 <- data.frame(pca1$x, Species=iris.repeated2$Species)
head(pca_data1)# iris.repeated2

pca_data2 <- data.frame(pca2$x, Species=iris.repeated3$Species)
head(pca_data2)# iris.repeated3
```
Now we can plot as usual:
```{r}
ggplot(pca_data, aes(x=PC1, y=PC2, color=Species)) + geom_point()
ggplot(pca_data1, aes(x=PC1, y=PC2, color=Species)) + geom_point()# iris.repeated2
ggplot(pca_data2, aes(x=PC1, y=PC2, color=Species)) + geom_point()# iris.repeated2
```
In the PC2 vs PC1 plot, versicolor and virginica are much better separated.

Next, let's look at the rotation matrx:
```{r}
pca$rotation
pca1$rotation# iris.repeated2
pca2$rotation# iris.repeated3
```
It tells us how much each variable contributes to each principal component. For example, `Sepal.Width` contributes little to PC1 but makes up much of PC2. Often it is helpful to plot the rotation matrix as arrows. This can be done as follows:
```{r}
# capture the rotation matrix in a data frame
rotation_data <- data.frame(pca$rotation, variable=row.names(pca$rotation))
# define a pleasing arrow style
arrow_style <- arrow(length = unit(0.05, "inches"),
                     type = "closed")
# now plot, using geom_segment() for arrows and geom_text for labels
ggplot(rotation_data) + 
  geom_segment(aes(xend=PC1, yend=PC2), x=0, y=0, arrow=arrow_style) + 
  geom_text(aes(x=PC1, y=PC2, label=variable), hjust=0, size=3, color='red') + 
  xlim(-1.,1.25) + 
  ylim(-1.,1.) +
  coord_fixed() # fix aspect ratio to 1:1

# iris.repeated2
# capture the rotation matrix in a data frame
rotation_data1 <- data.frame(pca1$rotation, variable=row.names(pca1$rotation))
# now plot, using geom_segment() for arrows and geom_text for labels
ggplot(rotation_data1) + 
  geom_segment(aes(xend=PC1, yend=PC2), x=0, y=0, arrow=arrow_style) + 
  geom_text(aes(x=PC1, y=PC2, label=variable), hjust=0, size=3, color='red') + 
  xlim(-1.,1.25) + 
  ylim(-1.,1.) +
  coord_fixed() # fix aspect ratio to 1:1

# iris.repeated3
# capture the rotation matrix in a data frame
rotation_data1 <- data.frame(pca2$rotation, variable=row.names(pca1$rotation))
# now plot, using geom_segment() for arrows and geom_text for labels
ggplot(rotation_data1) + 
  geom_segment(aes(xend=PC1, yend=PC2), x=0, y=0, arrow=arrow_style) + 
  geom_text(aes(x=PC1, y=PC2, label=variable), hjust=0, size=3, color='red') + 
  xlim(-1.,1.25) + 
  ylim(-1.,1.) +
  coord_fixed() # fix aspect ratio to 1:1
```
We can now see clearly that `Petal.Length`, `Petal.Width`, and `Sepal.Length` all contribute to PC1, and `Sepal.Width` dominates PC2.


Finally, we want to look at the percent variance explained. The `prcomp()` function gives us standard deviations (stored in `pca$sdev`). To convert them into percent variance explained, we square them and then divide by the sum over all squared standard deviations:
```{r}
percent <- 100*pca$sdev^2/sum(pca$sdev^2)
percent

percent1 <- 100*pca1$sdev^2/sum(pca1$sdev^2)
percent1# iris.repeated2

percent2 <- 100*pca2$sdev^2/sum(pca2$sdev^2)
percent2# iris.repeated3
```
The first component explains 73% of the variance, the second 23%, the third 4% and the last 0.5%.
We can visualize these results nicely in a bar plot:
```{r}
perc_data <- data.frame(percent=percent, PC=1:length(percent))
ggplot(perc_data, aes(x=PC, y=percent)) + 
  geom_col() + 
  geom_text(aes(label=round(percent, 2)), size=4, vjust=-.5) + 
  ylim(0, 100)

perc_data1 <- data.frame(percent=percent1, PC=1:length(percent1))
ggplot(perc_data1, aes(x=PC, y=percent)) + 
  geom_col() + 
  geom_text(aes(label=round(percent, 2)), size=4, vjust=-.5) + 
  ylim(0, 100)# iris.repeated2

perc_data2 <- data.frame(percent=percent2, PC=1:length(percent2))
ggplot(perc_data2, aes(x=PC, y=percent)) + 
  geom_col() + 
  geom_text(aes(label=round(percent, 2)), size=4, vjust=-.5) + 
  ylim(0, 100)# iris.repeated3
```


*Conclusion: we do not perform PCA on repeated measurements, but will select one measurement for each lesion and then calculate the pca results for the rest of observations.*